Early Warning Signals and the Prosecutor's Fallacy

### 2012-05-16 fallacy.tex
fallacy updated

\begin{align}
  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master} \\
    b_n &= \frac{e K n^2}{n^2 + h^2} \\
    d_n &= e n + a
\end{align}

a model with a linear death rate and density dependent birth rate that drives
the Allee effect at low densities and limits growth at high densities.  The
key feature of this model is the alternate stable states introduced by this effect;
other functional forms for Eq.~\ref{master} could serve equally well for these
simulations~\citep[see \emph{e.g.}][]{Scheffer2001}. 
Though this system can be forced through a saddle-node bifurcation by
increasing the death rate, in these simulations all parameters
are held constant and no bifurcation occurs.  Consequently we do not 
anticipate an early warning signal of an approaching bifurcation.
The simulation starts from the positive equilibrium population size.
Though the chance of a transition across the Allee threshold in any 
given time step are small, given enough time this system will eventually
experience such a rare event driving the population extinct.  We evolved
each replicate 50,000 time steps, sampling the system every 50 time steps.  
In this time window 266 of the 1,000 replicates experience population collapse.  
To keep the examples of comperable sample size, we focus on a section of the data
500 time points prior to the system approaching the transition.  
Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
as an illustration.

@@

To test for the presence of a warning signal in these indicators we computed
values of Kendall's $\tau$ for both indicators for each of the 266 replicates.
Kendall's $tau$ takes values in $(-1, 1)$, and is frequently used to identify
an increasing trend ($\tau > 0 )$ in early warning signals.~\citep{Dakos2008, Dakos2011}.
indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
values observed across these replicates is shown in Figure~\ref{fig:indicator}.
More positive values indicate stronger trends. The common statistical significance
test used for Kendall's $tau$ assumes independence of the variables to achieve
a normal distribution, this isn't satisfied for indicators coming from a time 
series and computed with a sliding window.  For a more robust comparison, we 
compare the distribution of this statistic from all the simulations to the 
distribution conditioned on experiencing a chance transition to the alternative
stable state, that is, from the replicates which are known to crash.  To avoid 
an effect of sample size the timeseries are all chosen to be the same length.  
As these examples have been selected from a system that is not approaching a 
bifurcation, an indicator that does not suffer from the bias of Prosecutor's fallacy
should have the same distribution in both cases.   Annotated, parallelized code to 
replicate the analysis and produce the figures is provided in the appendix. 
(Currently at \href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=6in]{indicators.pdf}
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two early warning indicators (variance, autocorrelation) on replicates conditionally selected for having collapsed by chance in simulations. Positive values of $\tau$ correspond to a pattern of an indicator increasing with time; typically taken as evidence that a system is approaching a critical transition.  In these simulations, the pattern arises instead from the Prosecutor's fallacy of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

Historical examples of systems that have undergone sudden transitions seem
to offer an invaluable opportunity to test methods for the detection of
early warning signals that could have forseen the collapse.  However,
such data must be used with care, as the process of selecting systems
already known to have collapsed opens up the possibility that these
transitions were the result of rare chance events rather than examples
of the more deterministic tipping point behaviour we wish to detect.
When we think about forecasting potential collapse in a single system,
these purely-chance transitions may be rare enough to ignore.  When we
comb the historical record for any available data on sudden transitions,
our odds of finding these chance transitions may be much higher.

@@

The problem we highlight ultimately stems from the difficulty of having
only a single realization with which to examine a complex problem.
The only way to deal with this problem is through replication, as
can be done in an experimental system in laboratory manipulations
such as~\citet{Drake2010, Veraart2011} and at the scale of whole lake
ecosystems in~\citet{Carpenter2011}.  Historical data will not suffice,
more such replicated experiments are needed to test these patterns.

 \section{Acknowledgements}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308 to CB. 
 \section{References}%bibliography
 \bibliography{fallacy}
 \end{document}

>>>>

\begin{align}
  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master} \\
    b_n &= \frac{e K n^2}{n^2 + h^2} \\
    d_n &= e n + a
\end{align}
a model with a linear death rate and density dependent birth rate
that drives the Allee effect at low densities and limits growth
at high densities.  The key feature of this model is the alternate
stable states introduced by this effect; other functional forms for
Eq.~\ref{master} could serve equally well for these simulations~\citep[see
\emph{e.g.}][]{Scheffer2001}.  Though this system can be forced
through a saddle-node bifurcation by increasing the death rate, in
these simulations all parameters are held constant and no bifurcation
occurs.  Consequently we do not anticipate an early warning signal of
an approaching bifurcation.  The simulation starts from the positive
equilibrium population size.  Though the chance of a transition across
the Allee threshold in any given time step are small, given enough
time this system will eventually experience such a rare event driving
the population extinct.  We evolved each replicate 50,000 time steps,
sampling the system every 50 time steps.  In this time window 266 of the
1,000 replicates experience population collapse.  To keep the examples of
comparable sample size, we focus on a section of the data 500 time points
prior to the system approaching the transition.  Nine of these replicates
can be seen in Figure~\ref{fig:replicate_crashes}, as an illustration.

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $tau$ takes values in $(-1, 1)$, and is
frequently used to identify an increasing trend ($\tau > 0 )$ in early
warning signals.~\citep{Dakos2008, Dakos2011}.  The distribution
of $\tau$ values observed across these replicates is shown in
Figure~\ref{fig:indicator}.  More positive values indicate stronger
trends. The common statistical significance test used for Kendall's $tau$
assumes independence of the variables to achieve a normal distribution,
this isn't satisfied for indicators coming from a time series and
computed with a sliding window.  For a more robust comparison, we
compare the distribution of this statistic from all the simulations to
the distribution conditioned on experiencing a chance transition to the
alternative stable state, that is, from the replicates which are known
to crash.  To avoid an effect of sample size the time series are all chosen
to be the same length.  As these examples have been selected from a system
that is not approaching a bifurcation, an indicator that does not suffer
from the bias of Prosecutor's fallacy should have the same distribution
in both cases.   Annotated, parallelized code to replicate the analysis
and produce the figures is provided in the appendix.  (Currently at
\href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).

\begin{figure}
  \begin{center}
    \includegraphics[width=6in]{indicators.pdf}
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in gray bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

Historical examples of systems that have undergone sudden transitions seem
to offer an invaluable opportunity to test methods for the detection of
early warning signals that could have foreseen the collapse.  However,
such data must be used with care, as the process of selecting systems
already known to have collapsed opens up the possibility that these
transitions were the result of rare chance events rather than examples
of the more deterministic tipping point behaviour we wish to detect.
When we think about forecasting potential collapse in a single system,
these purely-chance transitions may be rare enough to ignore.  When we
comb the historical record for any available data on sudden transitions,
our odds of finding these chance transitions may be much higher.

@@

The problem we highlight ultimately stems from the difficulty of having
only a single realization with which to examine a complex problem.
The only way to deal with this problem is through replication, as
can be done in an experimental system in laboratory manipulations
such as~\citet{Drake2010, Veraart2011} and at the scale of whole lake
ecosystems in~\citet{Carpenter2011}.  Historical data will not suffice,
more such replicated experiments are needed to test these patterns.
Experimental procedures avoid the hazard of the Prosecutor's fallacy by
generating a complete sample of replicates, rather than selecting a subset
of cases some larger historical sample where the risk that if selecting a
chance transition cannot be ignored.  Further, we note that the influence
of the Prosecutors Fallacy need not be restricted to the context of early
warning signals, but may arise any time we search historical data for
examples containing some feature of interest and ignore the statistical
effect that selection process introduces in our data.

 \section{Acknowledgements}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department of
Energy grant DE-FG02-97ER25308 to CB.

 \section{References}%bibliography
 \bibliography{fallacy}
 \end{document}
